---
title: "project_demo"
author: "Peter Tran"
date: "2024-04-23"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(bayesrules)
library(bayesplot)
library(tidybayes)
library(rstanarm)
library(tidyverse)
library(rstan)
library(brms)     
library(tidyr)
library(patchwork)
library(broom.mixed)
```

## Dataset Description

The Abalone dataset:

Collected from 1994 from Tasmania Coast, where the abalone species is indicate as Haliotis species

Contained 9 prediction features. Obtained the actual age (in years) by adding 1.5 to the rings parameter


```{r}
# read the dataset into a data frame
data <- read.csv("http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data", header=FALSE)

# add column names
names(data) <- c("sex", "length", "diameter", "height", "weight.whole",
    "weight.shucked", "weight.viscera", "weight.shell", "rings")

# Add a new column 'Age' to the data by adding 1.5 to the 'Rings' column 
# This represent the actual abalone years
data$age <- data$rings + 1.5

head(data)


```

## Data Exploration

```{r}
# Function to generate plot
generate_plot <- function(data, x_var) {
  data %>%
    ggplot(aes_string(x = x_var, y = "age",color="sex")) +
    geom_point(data = data, size = 0.75)+
    labs(x = x_var, y = "Age")   # Adjust y-axis label as needed
}

plot_length <- generate_plot(data, "length")
plot_diameter <- generate_plot(data,"diameter")
plot_height <- generate_plot(data, "height")
plot_whole <- generate_plot(data, "weight.whole")
plot_shucked <- generate_plot(data, "weight.shucked")
plot_viscera <- generate_plot(data, "weight.viscera")
plot_shell <- generate_plot(data, "weight.shell")

# Arrange plots side by side
combined_plots <- plot_length + plot_diameter + plot_height + plot_whole + plot_shucked + plot_viscera + plot_shell

combined_plots
```

## Prior

Examined the mean age of Haliotis species type, in which typical age of abalone ranges from 10 years olds, and able to live up to at most 30-40 years old. We will denote our prior understanding with $N(10,5)$

```{r}
plot_normal(10,5)
```



## Regression Model 

```{r}
abalone_model_length <- stan_glm(age ~ length, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_diameter <- stan_glm(age ~ diameter, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)

```


```{r}
abalone_model_height <- stan_glm(age ~ height, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_whole <- stan_glm(age ~ weight.whole, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_shucked <- stan_glm(age ~ weight.shucked, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_viscera <- stan_glm(age ~ weight.viscera, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_viscera <- stan_glm(age ~ weight.viscera, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

```{r}
abalone_model_shell <- stan_glm(age ~ weight.shell, data = data,
                       family = gaussian,
                       prior_intercept = normal(10, 5),
                       chains = 4, iter = 5000*2, seed = 84735,
                       refresh = FALSE)
```

## Credible Intervals

```{r}
##################
# Tidy CI
##################
vars <- c("length", "diameter", "height", "weight.whole", "weight.shucked", "weight.viscera", "weight.shell")

model <- list(abalone_model_length, abalone_model_diameter, abalone_model_height, abalone_model_whole, abalone_model_shucked, abalone_model_viscera, abalone_model_shell)

estimate = c()
se = c()
conf_low = c()
conf_high = c()

for (i in seq(1:7)){
  
  ci <- tidy(model[[i]], effects = "fixed", 
             conf.int = TRUE, conf.level = 0.95)
  
  estimate <- append(estimate, ci[2,]$estimate)
  se <- append(se, ci[2,]$std.error)
  conf_low <- append(conf_low, ci[2,]$conf.low)
  conf_high <- append(conf_high, ci[2,]$conf.high)
}
df2 <- data.frame(vars, estimate, se, conf_low, conf_high)
df2_sorted <- df2 %>%
  arrange(desc(estimate))

# Display the sorted data frame
df2_sorted
```

## Mean Absolute Error (MAE)
```{r}
{r}
model_length_summary <- prediction_summary(abalone_model_length, data=data)
model_diameter_summary <- prediction_summary(abalone_model_diameter, data=data)
model_height_summary <- prediction_summary(abalone_model_height, data=data)
model_wholewgt_summary <- prediction_summary(abalone_model_whole, data=data)
model_shuckedwgt_summary <- prediction_summary(abalone_model_shucked, data=data)
model_viscerawgt_summary <- prediction_summary(abalone_model_viscera, data=data)
model_shellwgt_summary <- prediction_summary(abalone_model_shell, data=data)


#################################
# Mae prediction summary
#################################
vars <- c("length", "diameter", "height", "weight.whole", "weight.shucked", "weight.viscera", "weight.shell")
# i.e. model_length_summary <- prediction_summary(modelx, data=data)
# Where modelx is the respective bayesian model with a single predictor
model <- list(model_length_summary, model_diameter_summary, model_height_summary, model_wholewgt_summary, model_shuckedwgt_summary, model_viscerawgt_summary, model_shellwgt_summary)

mae = c()
mae_scaled = c()
within50 = c()
within95 = c()

for (i in seq(1:7)){
  mae <- append(mae, model[[i]]$mae)
  mae_scaled <- append(mae_scaled, model[[i]]$mae_scaled)
  within50 <- append(within50, model[[i]]$within_50)
  within95 <- append(within95, model[[i]]$within_95)
}
df <- data.frame(vars, mae, mae_scaled, within50, within95)
ranked_df <- df[order(df$mae_scaled), ]
ranked_df
```



